Self Referencing Sequences

### 2014-11-06 draft.tex
edited first half of equivalence section

Thus, the sequence is uniquely given by first ``writing down" $B_1$, followed by $B_2$, followed by $B_3$, etc. One detail is that it remains to check that $a_i$ is known when $B_i$ is written down. However, writing down a block always involves writing down at least one element of the sequence, so the value of $a_i$ is always known when $B_i$ is written down. Note that there is an edge case where $a_i$ is not yet written down if the starting number is 1 and $i =2$. However, in this case, the value of $a_i$ is still known (as it is given by the rule) and thus $B_i$ can still be written down.
\end{proof}
%
%\begin{lemma}
%Let $C$ be a generating cycle. Suppose $\{a_i\}$ and $\{b_i\}$ are both self-referencing sequences with $a_j \neq b_j$ and $j$ is the smallest such number, then $j = 0$.
%\end{lemma}
%
%\begin{proof}
%Suppose $j > 0$, then $a_{j-1} = b_{j-1}$ since $j$ was assumed to be the smallest index where $\{a_i\}$ and $\{b_i\}$ are different. Suppose $a_{j-1} \neq a_j$ and $b_{j-1} \neq b_j$, then we can ask what comes directly after $a_{j-1}$ is the generating cycle; by looking at sequence $\{a_i\}$, we get that $a_j$ comes after $a_{j-1}$, but by looking at the sequence $\{b_i\}$, we get that $b_j$ comes after $a_{j-1}$. This is a contradiction since $a_j \neq b_j$. So lets assume without loss of generality that $a_{j-1} = a_j$, and also, we will assume that $a_{j-1}$ belongs to the $k$th block. Since the sequences are the same up to index $j$, $b_{j-1}$ belongs to the $k$th block. However, this means that $a_k \geq b_k + 1$. So $k \geq j$ and $k$ designates the block length of the block containing $a_j$, and so $k\leq j$, so $k=j$. Figure~\ref{fig:unequalsequences} gives a visual representation of these sequences.
%
%\begin{figure}{hb}
%\label{fig:unequalsequences}
%\center
%\includegraphics[width=0.5\linewidth]{lemma1.pdf}
%\caption{Self-referencing sequences $a$ and $b$ which differ first at the $j$th element.}
%\end{figure}
%
%This implies that every block before the $k$th block just contains 1's since each block refered it itself. This means that either $j = 2$, or the sequence is not self-referential. But if $j=2$ and $a_1 = b_1 = 1$, then $a_2 \neq b_2$, which means both $a_2$ and $b_2$ comes after $1$ in the generating cycle. 
%\end{proof}
%
%
%
%We present an algorithm that produces the unique self-referencing sequence given a generating cycle and a starting element. Assume the generating cycle is $(a_1 a_2 \dots a_m)$ and that the starting element is some $a_i$. For notational simplicity, if we ever write down $a_k$ where $k > m$, we set this equal to $a_{k (\text{mod } m)}$.
%\begin{quote}
%\textbf{Algorithm:}\\
%Suppose the first number of the sequence is $a_i$. Then the first block of the sequence must be a block of $a_i$ instances of $a_i$, so we extend the sequence to this block. If $a_i$ is not 1, then there is a second element of that block. We read that second element, and then we add a block of $a_i$ instances of $a_{i+1}$ to the end of that sequence. If $a_i$ is equal to 1, we write down $a_{i+1}$ in the second spot, and use the value of $a_{i+1}$ to dictate the length of the second block.\\
%
%This procedure continues deterministically so long as there is an element to read. Every time we write down a block, we read 1 element and write down at least 1 element, thus the end of the sequence will always be at or after the spot we are reading.
%\end{quote}

@@

In this section, we consider long term equivalence of sequences. We introduce a formal definition of this and establish results that allow us to determine when two sequences are equivalent.

We would first like to make a formal definition of equivalence. 
\begin{definition}
We say that two sequence $\{a_i\}$ and $\{b_i\}$ are equivalent if there exists some integers $n, k \geq 0$ such that 
\begin{align*}
a_{n+i} = b_{k+i}, \quad i=1,2,3,\ldots 
\end{align*}
\end{definition}
Note that two sequences can only be equivalent if they are generated by the same cycle.

As a basic example of equivalence let us consider sequences generated by $C = (123)$. The sequence which starts with a 2 is 
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ ,
\end{align*}
and the sequence that starts with a 3 is 
\begin{align*}
3,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ .
\end{align*}
If we delete the first two numbers of the sequence starting with 2, and the first number of the sequence starting with 3, we see 
\begin{align*}
2,2,&|3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots\\
3,&|3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ ,
\end{align*}
that the resultant sequences appear to be identical and so the sequences should be equivalent. At an intuitive level it seems obvious that these two truncated sequences are identical as they begin with the same two numbers and the numbers which were removed do not refer to blocks to the right of these numbers, but we cannot actually show that these two sequences are equivalent using just our definition as we would need to check equality for an infinite number of terms.

In order to develop a method to address this problem we introduce the notion of a read distance.  
\begin{definition} Consider a sequence $\{a_i\}$. We define the read distance for an index $i$ as the number of terms in between it $a_i$ and the beginning of the $i$th block of numbers.
\end{definition}
The read distance can be any nonnegative integer. For example if we again consider generated by $C = (1 2 3)$ that starts with a $2$
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ ,
\end{align*}
the read distance for the first number, $2$, is $0$ since in a self-referencing sequence, the first number must refer to the first block. The read distance for the second $2$ is $1$ as the second block of numbers starts at the third number. We can continue to calculate the read distances for this sequence and see that they form a monotonically increasing sequence $0 ,1,2,4,6,6,6,6,7,8,9,\ldots$.

This monotonicity is true in general, and we can in fact make a stronger statement on the value of the read distance at a given index.
\begin{lemma} 
\label{readdistformula}
Let $\{a_i\}$ be any self referencing sequence generated by some cycle of positive integers. Then the read distance for the number at index $i>1$, is 
\begin{equation*}
\sum_{k=1}^{i-1}( a_k - 1).
\end{equation*} 
The read distance for the number at index $1$ is always 0.
\end{lemma}
\begin{proof}
We will prove this using induction. We know that for $i=1$ the read distance is $0$, as the first block always starts at the first number. When $i=2$ the read distance is $a_1-1$. The first block will be of length $a_1$ since the sequence is self referencing, therefore the second block will begin at position $a_1$. The number of elements between position $a_1$ and $2$ is exactly $a_1 -1$.

Now if we assume that our result holds for $i=l$ we would like to show that it holds for $i=l+1$. We know that the distance between the start of the $l$th block and position $l$ is 
\begin{equation*}
\sum_{k=1}^{l-1}( a_k - 1).
\end{equation*} 
This $l$th block has length $a_l$ and so the $(l+1)$th block will begin $a_l$ positions to the right of the beginning of the $l$th block.  Additionally $a_{l+1}$ is one position to the right of $a_l$. Therefore the beginning of the $(l+1)$th block is 
\begin{equation*}
(a_l-1 )+ \sum_{k=1}^{l-1} (a_k-1) = \sum_{k=1}^{l} (a_k-1)
\end{equation*} 
positions to the right of $a_{l+1}$. But this is exactly the desired read distance for $a_{l+1}$ and so we have our result.
\end{proof}
Note that because $\{a_i\}$ is generated by positive integers we always have $a_k-1\geq 0$ and so read distance is increases monotonically.

@@

Now that we have a nice expression for the value of the read distance at a given index we can show our first major result. It will allow us to check only finitely many indices in order to establish that two sequences are equivalent, thus solving the problem we encountered in our example.
\begin{theorem} 
\label{equivsuff}
Let $\{a_i\}, \{b_i\}$ be two self-referencing sequences generated by the same cycle. If there are integers $m,l$ such that the read distance for $a_m$ and $b_l$ is $d$ and 
\begin{equation*}
a_{m+k} = b_{l+k}, \quad 0 \leq k \leq d-1
\end{equation*}
then $\{a_i\}$ and $\{b_i\}$ are equivalent. 
\end{theorem}

\begin{proof}
We will proceed by contradiction so assume otherwise. That is there exists some $j \geq d$ such that 
\begin{equation*}
a_{m+j} \neq b_{l+j}
\end{equation*}
but 
\begin{equation*}
a_{m+k} = b_{l+k}
\end{equation*} 
for all $0 \leq k < j$.

Since the two sequences are both generated by the same cycle this implies that either $a_{m+j}$ or $b_{l+j}$ is part of the same block as the element proceeding it, while the other is not.
Without loss of generality we will assume that $a_{m+j}=a_{m+j-1}$ and $b_{l+j} \neq b_{l+j-1}$.

>>>>

Thus, the sequence is uniquely given by first ``writing down" $B_1$, followed by $B_2$, followed by $B_3$, etc. One detail is that it remains to check that $a_i$ is known when $B_i$ is written down. However, writing down a block always involves writing down at least one element of the sequence, so the value of $a_i$ is always known when $B_i$ is written down. Note that there is an edge case where $a_i$ is not yet written down if the starting number is 1 and $i =2$. However, in this case, the value of $a_i$ is still known (as it is given by the rule) and thus $B_i$ can still be written down.
\end{proof}

@@

In this section, we consider equivalence of the limiting behavior of sequences. We introduce a formal definition of this and establish results that allow us to determine when two sequences are equivalent.

We first make a formal definition of equivalence. 
\begin{definition}
We say that two sequences $\{a_i\}$ and $\{b_i\}$ are equivalent if there exist integers $n, k \geq 0$ such that 
\begin{align*}
a_{n+i} = b_{k+i}, \quad i=1,2,3,\ldots \ .
\end{align*}
\end{definition}
Note that two sequences can only be equivalent if they are generated by the same cycle.

As a basic example of equivalence, consider sequences generated by $C = (123)$. The sequence which starts with a 2 is 
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ ,
\end{align*}
and the sequence that starts with a 3 is 
\begin{align*}
3,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ .
\end{align*}
If we delete the first two numbers of the sequence starting with 2, and the first number of the sequence starting with 3, as follows
\begin{align*}
2,2,&|3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots\\
3,&|3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ ,
\end{align*}
we see that the resultant sequences are the same, and thus the original sequences are in some sense equivalent. At an intuitive level, it seems obvious that these two truncated sequences are identical since they begin with the same two numbers and the numbers which were removed do not refer to blocks to the right of these numbers. We now formalize this intuition for claiming that two sequences are in fact equivalent.

To do this, we first introduce the \emph{read distance}.
\begin{definition} Consider a sequence $\{a_i\}$. The read distance for an index $i$ is the difference $index(B_i) - i$, where $index(B_i)$ is the index of the first number in the $i$th block $B_i$ of numbers.
\end{definition}
Note that the read distance can be any nonnegative integer. For example if we again consider the sequence generated by $C = (1 2 3)$ that starts with a $2$
\begin{align*}
2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots \ ,
\end{align*}
the read distance for index 1 is $index(B_1) - 1 = 1 - 1 = 0$. The read distance for index 2 is $index(B_2) - 2 = 3 - 2 = 1$. We can continue to calculate the read distances at each index of this sequence and we see that they form a monotonically increasing sequence 
\[ 0,1,2,4,6,6,6,6,7,8,9,\ldots \]

This monotonicity is present in sequences other than ones generated by $(123)$. We can explain this monotonicity phenomenon with the following proposition.
\begin{proposition} 
\label{readdistformula}
Let $\{a_i\}$ be any self-referencing sequence generated by some cycle of positive integers. Then the read distance for index $j$ is 0 if $j =1$, and when $j>1$, the read distance is
\begin{equation*}
\sum_{k=1}^{j-1}( a_k - 1).
\end{equation*} 
\end{proposition}
\begin{proof}
We prove this using induction. We know that for $i=1$, the read distance is $0$ as the first block always starts at the first number. When $i=2$, the read distance is $a_1-1$. The first block will be of length $a_1$ since the sequence is self referencing, therefore the second block will begin at index $a_1+1$. So the read distance is $index(B_2) - 2 = a_1 -1$.

For the inductive step, suppose the formula holds for $i=l > 1$. This means that the read distance of index $l$ is
\begin{equation*}
index(B_l) - l = \sum_{k=1}^{l-1}( a_k - 1).
\end{equation*} 
The $l$th block has length $a_l$, which implies that $index(B_{l+1}) - index(B_l) = a_l$. Thus, the read distance of index $l+1$ is
\begin{align*}
index(B_{l+1}) - (l+1) = a_l - 1 + \sum_{k=1}^{l-1} (a_k-1) = \sum_{k=1}^{l} (a_k-1),
\end{align*}
which compeltes the induction.
\end{proof}
Note that because $\{a_i\}$ is generated by positive integers, we always have $a_k-1\geq 0$ and so the read distance increases monotonically.

@@

Now that we have a closed-form expression for the read distance at a given index, we prove our first major result: we can check only finitely many indices in order to establish that two sequences are equivalent.
\begin{theorem} 
\label{equivsuff}
Let $\{a_i\}, \{b_i\}$ be two self-referencing sequences generated by the same cycle. If there exist integers $m,l$ such that the read distances for $a_m$ and $b_l$ are both equal to $d$ and 
\begin{equation*}
a_{m+k} = b_{l+k}, \quad 0 \leq k \leq d-1,
\end{equation*}
then $\{a_i\}$ and $\{b_i\}$ are equivalent. 
\end{theorem}

\begin{proof}
We will proceed by contradiction, so first assume otherwise. That is, assume there exists some $j \geq d$ such that 
\begin{equation*}
a_{m+j} \neq b_{l+j}
\end{equation*}
but 
\begin{equation*}
a_{m+k} = b_{l+k}
\end{equation*} 
for all $0 \leq k < j$.

Since the two sequences are both generated by the same cycle, this implies that either $a_{m+j}$ or $b_{l+j}$ is part of the same block as the element proceeding it, while the other is not.
Without loss of generality we will assume that $a_{m+j}=a_{m+j-1}$ and $b_{l+j} \neq b_{l+j-1}$.
